---
title: "Final Project Presentation"
author: "Ethan Blauw"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(randomForest)
library(caret)
library(tidyverse)
library(dplyr)
library(rpart)
library(yardstick)
library(partykit)
library(tidymodels)
library(broom)
library(rattle)
library(rpart.plot)
library(RColorBrewer)
song_data <- read.csv("spotify_songs.csv")
```

## Intro and Hypotheses

I chose to do my project on a dataset containing around 30,000 songs from spotify with corresponding data.

My research questions are: Can we predict the popularity of a specific song from the variables provided above? Can we predict the genre of the song from the variable above?

My hypothesis for these questions are: We can predict the popularity within (hopefully) a small margin of error (1 - 5 'popularity points'). We can predict the genre of the a song based on the variables above.

Start off with by looking at what data we have:

```{r}
head(song_data)
```

```{r echo=FALSE, warning=FALSE}
song_data <- song_data %>%
  distinct(track_id, .keep_all = T) %>%
  na.omit(song_data)


factors_vector <- c('0' = 'C', '1' = 'C#', '2' = 'D', '3' = 'D#', '4' = 'E', '5' = 'F', '6' = 'F#', '7' = 'G', '8' = 'G#', '9' = 'A', '10' = 'A#', '11' = 'B', '-1' = 'NaN')
song_data <- song_data %>%
  mutate(key = factors_vector[as.character(key)]) %>%
  mutate(key = as.factor(key))

song_data <- song_data %>%
  mutate(playlist_genre = as.factor(playlist_genre))

song_data <- song_data %>%
  mutate(duration_m = duration_ms / 60000)

song_data <- song_data %>%
  mutate(song_year = substr(track_album_release_date, start = 1, stop = 4)) %>%
  mutate(song_year = as.numeric(song_year))

song_data <- song_data %>%
  select(c(track_name, 
           track_artist, 
           track_popularity, 
           track_album_release_date, 
           playlist_genre,
           danceability, 
           energy, 
           key, 
           loudness, 
           mode, 
           speechiness, 
           acousticness, 
           instrumentalness, 
           liveness, 
           valence, 
           tempo, 
           duration_m, 
           song_year))
song_data <- song_data %>%
  mutate(song_rating = ifelse(track_popularity <= 25, "garbage", ifelse(track_popularity <= 50, "decent", ifelse(track_popularity <= 75, "good", "exceptional")))) %>%
  mutate(song_rating = as.factor(song_rating))
```

There is too much information for it all to be helpful, so lets clean some of it up. There is a lot of code here so I will just tell you what I did.

## Data Cleaning

First off, select the columns that we want. There are a lot, so instead I will tell you what I did not take. Those columns are track_id, track_album_id, playlist_id, and playlist_subgenre.

Next I cleaned up the data by changing some columns into factors such as key and playlist_genre, and removed all of the duplicate entries in the dataset. I also turned the key column into the actual music key notation for better understanding.

Lastly, I created extra columns, one called duration_m which is the duration of the song in minutes not milliseconds, song_rating which is a factor of 4 elements consisting of garbage, decent, good, and exceptional in increments of 25, and song_year, which is the year from the track_album_release_date column.

## Summarizing the Data

Now lets summarize the data.

```{r}
song_data %>%
  summarise(
    mean_popularity = mean(track_popularity),
    .by = key
  ) %>%
  arrange(desc(mean_popularity))
song_data %>%
  summarise(
    mean_popularity = mean(track_popularity),
    .by = playlist_genre
  ) %>%
  arrange(desc(mean_popularity))
song_data %>%
  summarize(
    count = n(),
    .by = key
  ) %>%
  arrange(desc(count))
```

## Plotting the Data

Now lets look at some plots for a better understanding of the data

```{r echo=FALSE, warning=FALSE}
ggplot(song_data, aes(x = key, fill = playlist_genre)) +
  geom_bar(position = "dodge", stat = "count") +
  ggtitle('Distribution of Key by Genre') +
  xlab('Key') +
  ylab('Count') +
  scale_alpha(range = c(0.3, 0.8))

ggplot(song_data, aes(x = song_year, y = track_popularity, color = playlist_genre, alpha = .2)) +
  geom_count() +
  ggtitle('Song Year vs. Track Popularity') +
  xlab('Song Year') +
  ylab('Track Popularity') +
  scale_x_continuous(breaks = seq(min(song_data$song_year), max(song_data$song_year), by = 10))

ggplot(song_data, aes(x = track_popularity, y = duration_m, color = song_rating, alpha = 0.2)) +
  geom_count() +
  ggtitle('Track Popularity vs. Duration') +
  xlab('Track Popularity') +
  ylab('Song Duration (m)')
```

## Research questions

After all of that great exploration, lets get back to the research questions. I want to try to predict the track_popularity variable using the data above. I did this by creating a random forest and training it on some training data, then testing it on both the test_data, and the whole dataset.

```{r}
sample_index <- sample(1:nrow(song_data), 0.8 * nrow(song_data))
train_data <- song_data[sample_index, ]
test_data <- song_data[-sample_index, ]

# Train a random forest model for regression
pop_model <- randomForest(track_popularity ~ danceability + energy + key + loudness + mode + speechiness + acousticness + instrumentalness + liveness + valence + tempo + duration_m, data = train_data, ntree = 100)

# Predict the track_popularity on the test set
predictions <- predict(pop_model, newdata = test_data)

# Print the Mean Squared Error (MSE)
mse <- mean((test_data$track_popularity - predictions)^2)
print(paste("Mean Squared Error:", mse))

```

Now what is the accuracy of our model, giving our selves 10 points of error:

```{r echo=FALSE, warning=FALSE}
new_pred <- test_data %>%
  select(track_popularity) %>%
  bind_cols(
    data.frame(new_predictions = predict(pop_model, newdata = test_data))
  )
threshold <- 10

# Calculate accuracy as the percentage of predictions within the threshold
accuracy <- mean(abs(new_pred$track_popularity - new_pred$new_predictions) <= threshold) * 100

# Print the accuracy
cat("Accuracy:", round(accuracy, 2), "%\n")
```

What about within 5 points:

```{r echo=FALSE, warning=FALSE}
new_pred <- test_data %>%
  select(track_popularity) %>%
  bind_cols(
    data.frame(new_predictions = predict(pop_model, newdata = test_data))
  )
threshold <- 5

# Calculate accuracy as the percentage of predictions within the threshold
accuracy <- mean(abs(new_pred$track_popularity - new_pred$new_predictions) <= threshold) * 100

# Print the accuracy
cat("Accuracy:", round(accuracy, 2), "%\n")
```

What about predicting song_rating?

```{r}
sample_index <- sample(1:nrow(song_data), 0.8 * nrow(song_data))
train_data <- song_data[sample_index, ]
test_data <- song_data[-sample_index, ]

# Train a random forest model for regression
rating_model <- randomForest(song_rating ~ danceability + energy + key + loudness + mode + speechiness + acousticness + instrumentalness + liveness + valence + tempo + duration_m, data = train_data, ntree = 100)

# Predict the track_popularity on the test set
predictions <- predict(rating_model, newdata = test_data)
confusion <- confusionMatrix(test_data$song_rating, predictions)
cat("Accuracy", round(confusion$overall["Accuracy"] * 100, 2), "%\n")
```

Lastly, what about the other question of predicting the genre? Well, lets do the same thing again, with less code to show this time.

```{r warning=FALSE}
sample_index <- sample(1:nrow(song_data), 0.8 * nrow(song_data))
train_data <- song_data[sample_index, ]
test_data <- song_data[-sample_index, ]

genre_model <- randomForest(playlist_genre ~ track_artist + track_popularity + danceability + energy + key + loudness + mode + speechiness + acousticness + instrumentalness + liveness + valence + tempo + duration_m + song_year, data = train_data, ntree = 100)

predictions <- predict(genre_model, newdata = test_data)
confusion_genre <- confusionMatrix(test_data$playlist_genre, predictions)
cat("Accuracy", round(confusion_genre$overall["Accuracy"] * 100, 2), "%\n")
```

What do these results mean? Well, considering that the popularity of a song is based on factors of each person listening to a song, or public awareness of the song, and many other features, predicting low on this is not surprising . On the other hand, it feels as though it should be able to predict the song_genre better given the aspects of the song in the dataset.

## Conclusion

I think I have accomplished at least my first research question to the best of my ability with the information I have, and although the prediction is not great, I can predict, with a certain level of accuracy, the genre of the playlist that the song came from.
